---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Environment

```{r results="hide"}
plat <- sessionInfo()["platform"]
osv <- sessionInfo()["running"]
rv <- R.version$version.string
library(knitr)
knitrv <- packageVersion("knitr")
library(dplyr, quietly=TRUE, warn.conflicts=FALSE)
dplyrv <- packageVersion("dplyr")
library(lattice)
latticev <- packageVersion("lattice")
```

Platform `r plat`

Operating System `r osv`

`r rv`

knitr version `r knitrv`

dplyr version `r dplyrv`

lattice version `r latticev`




## Loading and preprocessing the data

The [activity.zip](https://github.com/DarrelSchwarz/RepData_PeerAssessment1/commit/9abcd99566ad027022e4aafbffb5768f9ee11182#diff-d732846a7761c0c1ec91a3ba65791c60) file was downloaded from github on "2019-07-15 15:37:29 AEST".

The file was originally committed on Feb 12, 2014 by user **rdpeng** commit id **9abcd99566ad027022e4aafbffb5768f9ee11182**

Note: Ensure you have the activity.zip file in your working directory.

```{r}
act <- read.csv(unzip("./activity.zip","activity.csv"), header=TRUE, sep=",")
str(act)
```

The date column has a type of Factor so convert to date for easier processing.
```{r}
act$date <- as.Date(act$date,"%Y-%m-%d")
str(act)
```




## What is mean total number of steps taken per day?

```{r compute}
daytots <- aggregate(act["steps"], by=act["date"], sum, na.rm=TRUE)
daymean <- mean(daytots$step, na.rm=TRUE)
daymedian <- median(daytots$step, na.rm=TRUE)
daydiff <- daymedian - daymean
```

The Mean of the total number of steps taken per day is `r daymean`.

The Median of the total number of steps taken per day is `r daymedian`.

The difference between the Median and Mean is `r daydiff` steps.

```{r}
hist(daytots$steps, main="Total Steps Taken Per Day", 
     xlab="Steps", col="green", breaks=seq(0, 25000, by=1000))
abline(v=daymean, lty=6, col="red", lwd=2)
text(daymean, 10, formatC(daymean, digits=0, format="f"), col="red", cex=0.5, adj=c(-.1, -.5))
abline(v=daymedian, lty=6, col='blue',lwd=2)
text(daymedian, 10, formatC(daymedian, digits=0, format="f"), col="blue", cex=0.5, adj=c(-.1, -.5))
legend("topright", legend=c("Mean", "Median"), col=c("red", "blue"), lwd=c(2,2))
```





## What is the average daily activity pattern?

```{r}
act[12:14,]
```

The interval is in 5 min steps and is stored as an integer as hhmm so need to convert to time so that plots don't have gaps on the interval axis (ie. gaps between 55 and 100 for each hour).

```{r}
int <- aggregate(act['steps'],by=act['interval'],mean, na.rm=TRUE )
int$time <- formatC(int$interval, width=4, digits = 0, format = "d", flag=0)
int$time <- strptime(int$time,format='%H%M')
maxSteps <- max(int$steps) 
maxStepsInt <- int$interval[which(int$steps==maxSteps)]
maxStepsTm <- strftime(int$time[which(int$steps==maxSteps)],format='%H:%M')
int[12:14,]
plot(int$time, int$steps, type="l", col="blue",
     main="Average Daily Activity\n(5 min intervals)",
     xaxt = 'n', xlab="Time",
     ylab="Average Number of Steps")
r <- as.POSIXct(round(range(int$time), "hours"))
axis.POSIXct(1, at=seq(r[1], r[2], by="hour"), format="%H:%M")
abline(v=as.numeric(int$time[which(int$steps==maxSteps)]), lty=6, col='red',lwd=2)
text(as.numeric(int$time[which(int$steps==maxSteps)]), maxSteps,
     paste("Max Steps ",formatC(maxSteps, digits=0, format="f"), " occurred at ",maxStepsTm),
     col="red", cex=0.75, adj=c(-0.05, -0.1))
```

The 5-minute interval with the maximum average steps of `r maxSteps` is `r maxStepsTm` (interval `r maxStepsInt` )




## Imputing missing values

```{r}
sumna <- sum(is.na(act$steps))
sumnaperc <- sum(sumna/length(is.na(act$steps)))*100
```
There are `r sumna` NA's which is `r sumnaperc`% of the data.

As each weekday appears to have a different interval/step pattern, the missing (NA) step values were replaced with the average steps for that weekday/interval.

```{r}
act$daytype <- factor(weekdays(act$date)) 
weekDayIntervalMean <- aggregate(act['steps'],by=c(act["interval"],act["daytype"]),mean, na.rm=TRUE)
act2 <- left_join(act, weekDayIntervalMean, by = c("daytype", "interval")) %>% 
    mutate(steps = ifelse(is.na(steps.x), steps.y, steps.x)) %>% 
    select("date","interval", "daytype", "steps")
```


```{r}
daytots2 <- aggregate(act2['steps'],by=act['date'],sum)
daymean2 <- mean(daytots2$step, na.rm=TRUE)
daymedian2 <- median(daytots2$step, na.rm=TRUE)
daydiff2 <- daymedian2 - daymean2
```

The Mean of the total number of steps taken per day is `r daymean2` (previously `r daymean`).

The Median of the total number of steps taken per day is `r daymedian2` (previously `r daymedian`).

The difference between the Median and Mean is now `r daydiff2` steps (previously `r daydiff`).

```{r}

hist(daytots2$steps, main='Total Steps Taken Per Day(Imputed)', 
     xlab='Steps', col='green', breaks=seq(0,25000,by=1000))
abline(v=daymean2, lty=6, col="red", lwd=2)
text(daymean2, 10, formatC(daymean2, digits=0, format="f"), col="red", cex=0.5, adj=c(-.1, -.75))
abline(v=daymedian2, lty=6, col='blue',lwd=2)
text(daymedian2, 10, formatC(daymedian2, digits=0, format="f"), col="blue", cex=0.5, adj=c(-.1, 0.5))
legend("topright", legend=c("Mean", "Median"), col=c("red", "blue"), lwd=c(2,2))

```

By imputing the missing data, the number of days with zero count has dropped and the histogram looks more like a bell curve.

This has also resulted in the difference between the mean and median narrowing.




## Are there differences in activity patterns between weekdays and weekends?

Add a two level Factor "weekday" & "weekend". Then aggregate by Factor.

```{r}
act2$daytype <- factor(weekdays(act2$date) %in% c("Saturday","Sunday"), 
                       levels=c(FALSE,TRUE) ,labels=c("weekday","weekend"))
int <- aggregate(act2['steps'],by=c(act2["interval"],act2["daytype"]),mean, na.rm=TRUE )
```

The interval is in 5 min steps so convert to time again before plotting.

```{r}
int$time <- formatC(int$interval, width=4, digits = 0, format = "d", flag=0)
int$time <- strptime(int$time,format='%H%M')
r <- as.POSIXct(round(range(int$time), "hours"))
xyplot(int$steps~as.POSIXct(int$time)  | int$daytype, type="l",layout=c(1,2),
       main="Average Daily Activity (Imputed)\n(5 min intervals)",
       xlab="Time",
       ylab="Average Number of Steps",
       scales=list(x=list(at=seq(as.POSIXct(r[1]), by="3 hour", length=9),
                          labels=format(seq(as.POSIXct(r[1]), by="3 hour",
                                            length=9), "%H:%M"))))
```



